# ATAC-seq Analysis {.unnumbered}

**Work in progress**

## Filter raw fastq files

Use [fastp](https://github.com/OpenGene/fastp) to perform quality trimming
on raw fastq files. Turn on adapter detection for paired end reads.

```{bash}
#| eval: false

#!/usr/bin/env bash
#
# Run fastp on the raw fastq files
#
# ----------------------------------------------------------------------------
set -Eeou pipefail

FQ=/path/to/raw-fastq/00_fastq
SAMPLES=/path/to/sample-names.txt   
OUT=/path/to/01_fastp
THREADS=12

mkdir -p $OUT

for SAMPLE in $(cat $SAMPLES)
do
    fastp -i $FQ/${SAMPLE}_R1.fastq.gz \
          -I $FQ/${SAMPLE}_R2.fastq.gz \
          -o $OUT/${SAMPLE}.trimmed.1.fq.gz \
          -O $OUT/${SAMPLE}.trimmed.2.fq.gz \
          -h $OUT/${SAMPLE}.fastp.html \
          -j $OUT/${SAMPLE}.fastp.json \
	        --detect_adapter_for_pe \
          -w $THREADS
done
```

`sample-names.txt` is a plain text file listing the basenames for all samples. 
For example, if you have "sample1_R1.fq.gz", "sample2_R1.fq.gz", 
"sample3_R1.fq.gz" then `sample-names.txt` would be:

```
sample1
sample2
sample3
```

This file gets used throughout. 

## Perform alignment with Bowtie2

Use `--very-sensitive` mode and set max insertion size to 1,000. The output is
piped to `samtools fixmate` for adding mate tag information and then to 
`samtools sort -n` to create name sorted BAM files. 

It's assumed that you have a bowtie2 index generated for your species of 
interest. The number of jobs, threads, and memory allowed per thread will all
be machine dependent.

```{bash}
#| eval: false

#!/usr/bin/env bash
#
# Align trimmed reads with bowtie2, fixmate tags, and output name sorted BAMs
#
# ----------------------------------------------------------------------------
set -Eeou pipefail

FQ=/path/to/01_fastp
SAMPLES=/path/to/sample-names.txt   
OUT=/path/to/02_align
IDX=/path/to/bt2_idx
INSERT=1000
JOBS=6
THREADS=4

mkdir -p $OUT

parallel --jobs $JOBS \
  "bowtie2 -x $IDX \
    -1 $FQ/{}.trimmed.1.fq.gz \
    -2 $FQ/{}.trimmed.2.fq.gz \
    --very-sensitive \
    --threads $THREADS \
    --maxins $INSERT | \
    samtools fixmate -m -@$THREADS - - | \
    samtools sort -n -@$THREADS -m4G -o $OUT/{}.bam - " :::: $SAMPLES
```

## Call peaks with `Genrich`

[Genrich](https://github.com/jsh58/Genrich) is a fast peak caller designed to 
work with single samples or replicates. It has a dedicated option for peak 
calling ATAC-seq samples.

The options used below:

- Output pileups over called regions in a pseudo bedfile (-k)
- Output the log file (-f) which can be used to recall peaks with different -p and -a values
- Exclude regions in blacklist (-E)
- Exclude reads mapping to chrM (-e)
- Call in ATAC-seq mode (-j)
- Remove PCR duplicates (-r)
- Specify p-value threshold for peak calling (-p)
- Skip name sorting check (-S)

```{bash}
#| eval: false

#!/usr/bin/env bash
#
# Call ATAC-seq peaks using Genrich on the name sorted BAMs
#
# ----------------------------------------------------------------------------
set -Eeou pipefail

SAMPLES=/path/to/sample-names.txt   
BAM=/path/to/02_align
OUT=/path/to/03_callpeak
EXCLUDE=/path/to/excluderanges.bed
PVAL=0.05
JOBS=6

parallel --jobs $JOBS \
    "Genrich -t $BAM/{}.bam \
        -o $OUT/{}.narrowPeak \
        -k $OUT/{}.pileup.txt \
        -f $OUT/{}.log \
        -E $EXCLUDE \
        -p $PVAL -j -r -e chrM -S " :::: $SAMPLES
```

### Exclusion lists

The “excluderanges.bed” file used above contains blacklisted regions of 
high-mappability that can be excluded from peak calling. This BED file can be 
created using the [excluderanges](https://www.bioconductor.org/packages/release/data/annotation/vignettes/excluderanges/inst/doc/excluderanges.html) R package. For example, to generate the 
excluded regions for mm39:

```{r}
#| eval: false

suppressMessages(library(GenomicRanges))
suppressMessages(library(AnnotationHub))

ah <- AnnotationHub()
query_data <- subset(ah, preparerclass == "excluderanges")
mm39_exclude_gr <- query_data[["AH107321"]]
rtracklayer::export.bed(mm39_exclude_gr, "mm39-excluderanges.bed")
```

## Determine consensus peaks for replicate samples

After peak calling finishes, I create consensus peak calls for replicate 
samples. The consensus peak calls can be used in downstream differential 
accessibility analysis or for visualization. Creating consensus peak calls
can be done in R by requiring all or some peaks to be called across replicate
samples.

```{r}
#| eval: false
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(GenomicRanges))


# Read in peak calls
peak_files <- list.files(here("data", "04_callpeak"), pattern="*.narrowPeak", full.names=TRUE)
names(peak_files) <- gsub(".narrowPeak", "", basename(peak_files))
peak_calls <- lapply(peak_files, rtracklayer::import)

# Separate into groups 
control_calls <- peak_calls[c("control1", "control2", "control3")]
treatment_calls <- peak_calls[c("treatment1", "treatment2", "treatment3")]
control_calls <- GRangesList(control_calls)
treatment_calls <- GRangesList(treatment_calls)

# Compute coverage across ranges
control_coverage <- coverage(control_calls)
treatment_coverage <- coverage(treatment_calls)

# Determine regions with coverage in N replicates -- here require all 3 to have the same peaks
control_covered <- GRanges(slice(control_coverage, lower=length(control_calls), rangesOnly=TRUE))
treatment_covered <- GRanges(slice(treatment_coverage, lower=length(treatment_calls), rangesOnly=TRUE))

# Close gaps in the resulting regions -- min.gapwidth can be adjusted 
control_consensus <- reduce(control_covered, min.gapwidth=101)
treatment_consensus <- reduce(treatment_covered, min.gapwidth=101)
all_consensus <- union(control_consensus, treatment_consensus)

# Export as BED files
rtracklayer::export.bed(control_consensus, here("data", "04_callpeak", "control-consensus.bed"))
rtracklayer::export.bed(treatment_consensus, here("data", "04_callpeak", "treatment-consensus.bed"))
rtracklayer::export.bed(all_consensus, here("data", "04_callpeak", "consensus.bed"))
```

## Create raw signal tracks

Genrich outputs a bedgraph-ish file using the `-k` flag that can be used to 
quickly create a a bigwig file of the raw signal which can be viewed in IGV. I 
find it useful to look at the raw signal tracks after peak calling so I can 
assess the absolute magnitude of the pileup that went into calling a peak. 
Then, if the peaks look too permissive or too stringent, adjust the peak calls.

Creating bigwigs from the bedgraph-ish files can be accomplished using `awk` 
and [bedGraphToBigWig](https://www.encodeproject.org/software/bedgraphtobigwig/)
from UCSC. 

```{bash}
#| eval: false

#!/usr/bin/env bash
#
# Convert the pileups computed by Genrich from bedGraph-ish files to bigwigs
#
# ----------------------------------------------------------------------------
set -Eeou pipefail

SAMPLES=/path/to/sample-names.txt   
PEAKS=/path/to/03_callpeak
OUT=/path/to/04_bg2bw
CHROM_SIZES=/path/to/chrom.sizes
JOBS=6

mkdir -p $OUT

echo "Extracting bedGraphs from experimental signal..."
parallel --jobs $JOBS \
  "awk 'NR > 2 { print \$1, \$2, \$3, \$4 }' $PEAKS/{}.pileup.txt | \
  LC_ALL=C sort -k1,1 -k2,2n > $OUT/{}.sorted.bedGraph" :::: $SAMPLES

echo "Creating bigwigs from pileups..."
parallel --jobs $JOBS "bedGraphToBigWig $OUT/{}.sorted.bedGraph $CHROM_SIZES $OUT/{}.bw" :::: $SAMPLES

echo "Cleaning up intermediate bedGraphs..."
find $OUT -type f -name "*.bedGraph" -delete
```

### Chrom sizes files

You can supply a URL to `bedGraphToBigWig` to get the chrom.sizes if you're 
genome is supported. Otherwise, you can create a chrom.sizes file using another 
UCSC utility, `faSize`, and your genome fasta.

```{bash}
#| eval: false
faSize -detailed -tab genome.fasta > genome.chrom.sizes
```

## ATACSeqQC

**TODO: complete this section**

The [ATACseqQC](https://bioconductor.org/packages/release/bioc/vignettes/ATACseqQC/inst/doc/ATACseqQC.html)
R package can be used to calculate various QC stats on the filtered BAM files. 
These should be assessed closely and compared to the 
[ENCODE data standards](https://www.encodeproject.org/atac-seq/).

## Differential abundance using `csaw` and `edgeR`

**TODO: elaborate on this section**

[csaw](https://bioconductor.org/books/release/csawBook/) can be used to perform
differential abundance over sliding windows *or* by counting reads in the 
consensus peaks defined above. This [paper](https://epigeneticsandchromatin.biomedcentral.com/articles/10.1186/s13072-020-00342-y) 
contains some code for performing and assessing ATAC-seq DA analysis using both 
methods. 

I usually generate read counts over the consensus peaks using `featureCounts` 
from the [Rsubread](https://bioconductor.org/packages/devel/bioc/vignettes/Rsubread/inst/doc/SubreadUsersGuide.pdf) package. An SAF formatted file can be generated easily from the 
"consensus.bed" file. To get the read counts to align with the Genrich called 
peaks, set the `-read2Pos 5` flag to count alignments over the 5' end of reads.

## Create TSS and region plots with `deeptools`

[deeptools](https://deeptools.readthedocs.io/en/latest/) can be used to compute
normalized coverage files in bigwig format for visualization in IGV. 
Normalization in `deeptools` can be really context dependent and what 
normalization strategy to use can often be unclear. 

If you've performed differential abundance analysis and have global scaling
normalization factors, they can be used at this step to generate [TMM normalized](https://www.biostars.org/p/413626/#414440) 
bigwigs. In the absence of scaling factors, I've seen people use RPGC 
normalization or CPM for ATAC-seq. 

Then you can run `deeptools` using the TSS regions of coding genes and called
consensus peaks. 

```{bash}
#| eval: false

#!/usr/bin/env bash
#
# Compute normalized coverage bigwigs for visualization
#
# 'deeptools' mamba env must be activated before running
# -----------------------------------------------------------------------------
SAMPLES=/path/to/sample-names.txt
BLACKLIST=/path/to/excluderanges.bed
CODING=/path/to/coding.bed
BAM=/path/to/02_align
PEAKS=/path/to/consensus.bed
OUT=/path/to/05_deeptools
GENOME_SIZE=2654621783  # mm39 specific value, see https://deeptools.readthedocs.io/en/latest/content/feature/effectiveGenomeSize.html
JOBS=6
THREADS=4

mkdir -p $OUT

echo "Position sorting BAMs..."
parallel --jobs $JOBS "samtools sort -@$THREADS -m4G -o $OUT/{}.sorted.bam $BAM/{}.bam" :::: $SAMPLES
parallel --jobs $JOBS "samtools index $OUT/{}.sorted.bam" :::: $SAMPLES

echo "Computing coverage over all BAM files..." 
parallel --jobs $JOBS \
    "bamCoverage --bam $BAM/{}.sorted.bam \
                 --outFileName $OUT/{}.bw \
                 --outFileFormat 'bigwig' \
                 --normalizeUsing 'RPGC' \
                 --effectiveGenomeSize $GENOME_SIZE \
                 --binSize 1 \
                 --extendReads \
                 --blackListFileName $BLACKLIST \
                 --numberOfProcessors $THREADS \
                 --ignoreDuplicates" :::: $SAMPLES

echo "Computing matrix over TSS regions..."
computeMatrix reference-point \
 --regionsFileName  \
 --scoreFileName $OUT/*.bw \
 --outFileName $OUT/tss.mat.gz \
 --referencePoint TSS \
 --beforeRegionStartLength 3000 \
 --afterRegionStartLength 3000 \
 --blackListFileName $BLACKLIST \
 --missingDataAsZero \
 --numberOfProcessors $JOBS

echo "Computing coverage over peak BEDs..."
computeMatrix reference-point \
  --regionsFileName $PEAKS \
  --scoreFileName $OUT/*.bw \
  --outFileName $OUT/peak.mat.gz \
  --referencePoint 'center' \
  --beforeRegionStartLength 3000 \
  --afterRegionStartLength 3000 \
  --blackListFileName $BLACKLIST \
  --missingDataAsZero \
  --numberOfProcessors $JOBS

echo "Plotting heatmap of TSS enrichment..."
plotHeatmap -m $OUT/tss.mat.gz \
  -o $OUT/tss-heatmap.pdf \
  --dpi 300 \
  --colorMap "viridis" \
  --boxAroundHeatmaps 'no' \
  --samplesLabel "Ctl 1" "Ctl 2" "Ctl 3" "Trt 1" "Trt 2" "Trt 3" \
  --regionsLabel "Protein Coding Genes" \
  --yAxisLabel "RPGC" \
  --perGroup \
  --heatmapHeight 12 \
  --heatmapWidth 8 \
  --plotFileFormat "pdf"

echo "Plotting heatmap of MACS peak enrichment..."
plotHeatmap -m $OUT/peak.mat.gz \
  -o $OUT/peak-heatmap.pdf \
  --dpi 300 \
  --colorMap "viridis" \
  --boxAroundHeatmaps 'no' \
  --samplesLabel "Ctl 1" "Ctl 2" "Ctl 3" "Trt 1" "Trt 2" "Trt 3" \
  --regionsLabel "Consensus Peak Calls" \
  --refPointLabel "Center of Peak" \
  --yAxisLabel "RPGC" \
  --perGroup \
  --heatmapHeight 12 \
  --heatmapWidth 8 \
  --plotFileFormat "pdf"
```

### Getting TSS coding region ranges

For ATAC-seq, I plot the enrichment over the TSS region for each sample as 
well as the coverage centered over the consensus peaks. To generate a BED file 
for protein coding gene regions ("coding.bed" above) in R:

```{r}
#| eval: false

# Get mouse annotation GTF, for example
url <- "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M37/gencode.vM37.annotation.gtf.gz"
gtf <- rtracklayer::import(url)

# Extract the protein coding gene ranges only
coding <- gtf[gtf$gene_type == "protein_coding" & gtf$type == "gene", ]

# rtracklayer complains if score is NA or non-numeric
coding$score <- 1L

# Export to a BED file for use in deeptools
rtracklayer::export.bed(coding, here("doc", "gencode.vM37.coding.bed"))
```

## Motif analysis with `MEME`

The [MEME suite](https://meme-suite.org/meme/) can be used to perform motif 
analysis on peaks that have been determined to have differential abundance. 
You can use R to generate fasta files for these regions based on GRanges 
extracted from DA analysis. These fasta files can then be used as input to
`XSTREME`, which performs de novo motif discovery, enrichment analysis, and 
comparisons with known motif databases.

For example, you can extract fasta regions like so (this uses hg38 as an example):

```{r}
#| eval: false
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(Biostrings))
suppressPackageStartupMessages(library(BSgenome.Hsapiens.UCSC.hg38))
suppressPackageStartupMessages(library(GenomicRanges))


# Assuming you have a file from DA analysis that contains genomic coordinates
# of peaks and annotations - for example from ChIPseeker::annotatePeak()
peaks <- fread("annotated-peaks.tsv")
peaks <- makeGRangesFromDataFrame(peaks, keep.extra.columns = TRUE)

# Extract promoter peaks -- this can be any filtering you're interested in 
promoter_peaks <- peaks[peaks$annotation %like% "Promoter"]

# Extract and write out the hg38 sequences as fasta records
pmtr_seqs <- getSeq(BSgenome.Hsapiens.UCSC.hg38, promoter_peaks)
writeXStringSet(pmtr_seqs, filepath = "promoter-peaks.fasta", format = "fasta")
```

These fasta records can then be used in `XSTREME` analysis from the MEME suite. The meme formatted databases of known motifs can be downloaded from MEME's [website](https://meme-suite.org/meme/db/motifs)

```{bash}
#| eval: false

#!/usr/bin/env bash
#
# Run XSTREME motif analysis on promoter peak sequences
#
# -------------------------------------------------------------------------------------------------
FA=/path/to/promoter-peaks.fasta
DB=/path/to/HUMAN/HOCOMOCOv11_core_HUMAN_mono_meme_format.meme
OUT=/path/to/xstreme

xstreme --oc $OUT --p $FA --seed 123 --m $DB --no-pgc --dna
```
