# R programming basics {.unnumbered}

## Introduction to programming in R

There are tons of great resources for learning R. [R for Data Science](https://r4ds.had.co.nz/) is probably the most popular resource for new users to get up to speed with slicing and dicing data in R. The *R for Data Science* book, however, is taught from the perspective of the [Tidyverse](https://www.tidyverse.org/). The *Tidyverse* is an opinionated set of packages and functions that help users perform data manipulations primarily on `data.frame`s. While these packages and functions can be great for experienced users by providing ergonomic and consistent interfaces for `data.frame` manipulation, it is my personal belief that new users should first learn the base language, especially if their goal is to perform bioinformatics analysis. 

Bioinformatics tools rely heavily on subsetting and matrix manipulations. In my experience, users who start learning R using only functions from the *Tidyverse* have a difficult time understanding matrix manipulations and subsetting operations common in bioinformatics workflows. This becomes especially important when using [SummarizedExperiments](https://www.bioconductor.org/packages/devel/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html) - the backbone of many bioinformatics data structures in R. 

For this reason, we're going to focus on learning R from the ground up using functions that exist primarily in the base language. A great resource for learning base R quickly is Norm Matloff's *fasteR*, which can be found [here](https://github.com/matloff/fasteR). 

## Vectors

You may have only ever encountered R from the perspective of the [tidyverse](https://www.tidyverse.org/). `tidyverse` functions provide useful abstractions for munging [tidy data](https://vita.had.co.nz/papers/tidy-data.html); however, most genomics data is often best represented and operated on as matrices. Keeping your data in matrix format can provide many benefits as far as speed and code clarity, which in turn helps to ensure correctness. You can think of matrices as just fancy 2D versions of vectors. So what are vectors?

Vectors are the main building blocks of most R analyses and are the R objects that you'll interact with the most. Whenever you use the `c()` function ('concatenate'), like `x <- c('a', 'b', 'c')`, you're creating a vector. Vectors hold R objects and are the building blocks of more complex structures in R.

> **Note:** The following is heavily inspired by Norm Matloff's excellent [fasteR](https://github.com/matloff/fasteR) tutorial. Take a look there to get a brief and concise overview of base R. You should also check out the first few chapters of Hadley Wickham's amazing book [Advanced R](https://adv-r.hadley.nz/). The [first edition](http://adv-r.had.co.nz/) contains some more information on base R.

### Subsetting vectors

Below, we'll use the built-in R constant called `LETTERS`. The `LETTERS` vector is simply a list of all uppercase letters in the Roman alphabet.

```{r}
# "A" "B" "C", ...
LETTERS
```

We can subset the vector by position. For example, to get the 3rd letter we use the `[` operator and the position we want to extract. It is important to stop here and appreciate that R uses 1-based indexing. Meaning that positions in vectors start at the value of 1. This is different from other programming languages like Python that use [0-based indexing](https://en.wikipedia.org/wiki/Zero-based_numbering).

```{r}
# "A" "B" "C"
#  1   2   3
LETTERS[3]
```

We can also use a range of positions. The notation `3:7` is a shortcut that generates the numbers 3, 4, 5, 6, 7.

```{r}
LETTERS[3:7]
```

We don't have to select sequential elements either. We can extract elements by using another vector of positions.

```{r}
LETTERS[c(7, 5, 14, 14, 1, 18, 15)]
```

Vectors become really powerful when we start combining them with logical operations. R supports all of the usual logical and comparison operators you can expect from a programming language: `<`, `>`, `==`, `!=`, `<=`, `>=`, `%in%`, `&`, and `|`.

```{r}
my_favorite_letters <- c("A", "B", "C")

# See that this produces a logical vector of (TRUE/FALSE) values
# TRUE when LETTERS is one of my_favorite_letters and FALSE otherwise
LETTERS %in% my_favorite_letters

# We can use that same expression to filter the vector
LETTERS[LETTERS %in% my_favorite_letters]
```

You may have also noticed that we used another special operator above: `<-`. The `<-` operator is an assignment operator. It takes the object on the right-hand side and assigns it to the variable on the left-hand side. Assigning objects to variables is a basic concept across all programming languages. R actually has a [few ways](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/assignOps) to assign objects to variables, but using `<-` is by far the most common and preferred for a number of [reasons](https://stackoverflow.com/a/1742550).

This same kind of subsetting works on vectors that contain numeric data as well. For example, we can filter the measurements of annual flow of water through the Nile river like so:

*(Note: `Nile` is another built-in dataset)*

```{r}
# Any values strictly greater than 1200
Nile[Nile > 1200]

# Any even number - `%%` is the modulus operator (returns the remainder)
Nile[Nile %% 2 == 0]
```

At this point, it's important to take a step back and appreciate what R is doing. Each of the comparison operators that we used above is *vectorized*. This means that the comparison is applied to all elements of the vector at one time. If you're used to a programming language like Python, this might seem foreign at first. In Python, you would have to write a list comprehension to filter observations from a list that meet a certain condition (e.g., `[x for x in Nile if x > 1200]`). However, in R, most functions and operators are vectorized, allowing us to do things like `Nile > 1200` and have the comparison applied to all of the elements of the vector automatically.

### Subsetting data.frames

But these are just one-dimensional vectors. In R, we usually deal with `data.frame`s (tibbles for you tidyverse folks) and matrices. Lucky for us, the subsetting operations we learned for vectors work the same way for `data.frame`s and matrices.

Let's take a look at the built-in `ToothGrowth` dataset. The data consists of the length of odontoblasts in 60 guinea pigs receiving one of three levels of vitamin C by one of two delivery methods.

```{r}
head(ToothGrowth)
```

The dollar sign `$` is used to extract an individual column from the `data.frame`, which is just a vector.

```{r}
head(ToothGrowth$len)
```

We can also use `[[` to get the same thing. Double-brackets come in handy when your columns are not valid R names, since `$` only works when columns are valid names without spaces or special characters.

```{r}
head(ToothGrowth[["len"]])
```

When subsetting a `data.frame` in base R, the general scheme is:

```
df[the rows you want, the columns you want]
```

So in order to get the value at the 5th row of the first column we could do:

```{r}
ToothGrowth[5, 1]
```

Again, we can combine this kind of thinking to extract rows and columns matching logical conditions. For example, if we want to get all of the animals administered orange juice ('OJ'):

```{r}
ToothGrowth[ToothGrowth$supp == "OJ", ]
```

We can also combine logical statements. For example, to get all of the rows for animals administered orange juice *and* with an odontoblast length ('len') less than 10:

```{r}
ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$len < 10, ]

# We can also use the bracket notation to select specific rows and columns at the same time
# Although this gets a little difficult to read
ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$len < 10, c("len", "supp")]
```

It gets annoying typing `ToothGrowth` every time we want to subset the `data.frame`. Base R has a very useful function called `subset()` that can help us type less. `subset()` essentially 'looks inside' the `data.frame` for the given columns and evaluates the expression without having to explicitly tell R where to find them. Think of it like `dplyr::filter()`, if you are familiar with that function.

```{r}
subset(ToothGrowth, supp == "OJ" & len < 10)
```

### Subsetting Lists

Another data structure to be aware of, which is used frequently, is the List. We've actually already encountered lists above: `data.frame`s are really just lists where each vector contains the same data type and all list elements are the exact same length.

We can create a list in R using the `list()` function. Notice how each list element has a name and can contain a different type of data and number of data elements.

```{r}
l <- list(
  element1 = c(1, 10, 12, 3, 6, 12, 13, 2, 5, 6, 3, 7),
  element2 = c("a", "b", "c"),
  element3 = c(TRUE, TRUE, FALSE, FALSE, FALSE),
  element4 = c(0.001, 0.05, 0.86, 1.098, 345.0)
)
```

Lists can be tricky at first. To extract the underlying data from a particular list element, you can use `[[` or `$` (as in the case of `data.frame`s above). Like vectors, you can use either the index or the name of the element you wish to extract.

```{r}
l[[1]]

l[["element1"]]

l$element1
```

What is returned if you only use the single bracket `[`?

```{r}
l[1]
```

You get another List, but now with a single element. This behavior might seem unintuitive at first, but it can be very useful for creating new, smaller lists.

```{r}
numeric_l <- l[c(1, 4)]
```

### Subsetting matrices

Matrices behave much like `data.frame`s, but unlike `data.frame`s, matrices can only contain *one* type of data (e.g., all numeric or all character). This might sound like a limitation at first, but you'll soon come to realize that matrices are very powerful (and fast) to work with in R.

```{r}
set.seed(123)

# Create some random data that looks like methylation values
# runif() generates random numbers between 0 and 1
(m <- matrix(
  data = runif(6 * 10),
  ncol = 6,
  dimnames = list(
    paste0("CpG.", 1:10),
    paste0("Sample", 1:6)
  )
))
```

If we want to extract the value for CpG.3 for Sample3:

```{r}
m[3, 3]
```

Or all values of CpG.3 for every sample:

```{r}
m[3, ]

# Or refer to the row by its name
m["CpG.3", ]
```

Or all CpGs for Sample3:

```{r}
m[, 3]

# Or refer to the column by its name
m[, "Sample3"]
```

We can also apply a logical mask to the entire matrix at once. For example, the following will mark any value that is greater than 0.5 with `TRUE`.

```{r}
m > 0.5
```

We can use this kind of masking to filter rows of the matrix using some very helpful base R functions. For example, to get only those CpGs where 3 or more samples have a value > 0.5, we can use `rowSums()` like so:

```{r}
m[rowSums(m > 0.5) > 3, ]
```

This pattern is very common when dealing with sequencing data. Base R functions like `rowSums()` and `colMeans()` are specialized to operate over matrices and are the most efficient way to summarize matrix data. The R package [matrixStats](https://github.com/HenrikBengtsson/matrixStats) also contains highly optimized functions for operating on matrices.

Compare the elegance of the base R matrix subsetting above to the `tidy` solution given the same matrix:

```{r}
tidyr::as_tibble(m, rownames = "CpG") |>
  tidyr::pivot_longer(!CpG, names_to = "SampleName", values_to = "beta") |>
  dplyr::group_by(CpG) |>
  dplyr::mutate(n = sum(beta > 0.5)) |>
  dplyr::filter(n > 3) |>
  tidyr::pivot_wider(
    id_cols = CpG,
    names_from = "SampleName",
    values_from = "beta"
  ) |>
  tibble::column_to_rownames(var = "CpG") |>
  data.matrix()
```

*(Note: There's probably some kind of tidy solution using `across()` that I'm missing, but this is how most of the tidy code in the wild that I have seen looks).*

## Data Wrangling

Now that we've got a handle on some different R data types and how to slice and dice them, we can start learning basic data cleaning and exploratory data analysis. We'll focus on using `data.frame`s since they're the primary workhorse of data analysis in R. But remember, `data.frame`s are just lists of vectors with some special rules, so many concepts you learn will apply to `data.frame`s but also apply to vectors and lists.

### Data import

We'll use the built-in `penguins_raw` dataset to learn some basic data cleaning. This dataset is built into R version 4.5, so you can just load it by running `penguins_raw` if you have that R version installed. However, to illustrate data import, we'll read in the data from an external source.

The `read.csv()` function can read in comma-separated value files that are located either on your local machine or from remote sources if provided a URL.

```{r}
url <-
  "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/refs/heads/main/inst/extdata/penguins_raw.csv"

penguins <- read.csv(url)
```

The `read.csv()` function has many options for reading in data. If you want to learn about all of the options any particular R function has, you can prefix the function name with a `?` like, `?read.csv()` to bring up the help documentation.

### Data export

If you have a `data.frame` then you can also export it as a file for later. The most common format to export data to is a csv file like that used above. To export a csv file we can use the `write.csv()` function and supply the `data.frame` we wish to write out along with the filename we wish to write the data to. 

```{r}
write.csv(penguins, "~/Downloads/my-penguins.csv")
```

You can also save R objects into files that only R can read using the `saveRDS()` function. These objects can then be read back into an R session using the `readRDS()` function.  

### Viewing data

The code above read the data into a `data.frame` that we called `penguins`. We can take a look at the first few rows using the `head()` function.

```{r}
head(penguins)
```

If we want to get a general overview of the data types, we can use the `str()` (structure) function.

```{r}
str(penguins)
```

There are a few external packages that are also very useful for getting summaries of `data.frame`s. `Hmisc::describe()` and `skimr::skim()` are two standouts.

One of the most basic ways to get an idea of the data is to summarize each variable. There are a few functions we can use to get summaries of the data. The `table()` function will count the number of occurrences of each type in a vector.

For example, how many observations of each species of penguin are in the dataset?

```{r}
table(penguins$Species)
```

R also provides the typical summary functions that you would expect from a statistical programming language such as `mean()`, `median()`, `min()`, `max()`, and `length()`.

For example, what is the mean flipper length?

```{r}
mean(penguins$Flipper.Length..mm.)
```

Oh no! This returned `NA`, but there is clearly data in this column. What happened? Missing data is commonly observed across all data domains. `NA`s simply represent unknown values in this context, and it's impossible to know how to take the mean of a value that's known with a value that's unknown. It is for this reason that many R functions have an argument called `na.rm`. Setting `na.rm = TRUE` in these functions tells R to ignore the `NA` values.

```{r}
mean(penguins$Flipper.Length..mm., na.rm = TRUE)
```

Now we can see that the mean flipper length is ~200 mm across all observations in the dataset. Another extremely useful function is `summary()`. Running `summary()` on a numeric vector returns a lot of useful information at once.

```{r}
summary(penguins$Flipper.Length..mm., na.rm = TRUE)
```

Finally, if you are using an IDE like RStudio or Positron, you can run the `View()` function on your `data.frame`. This will bring up an interactive data viewer in a new tab.

### Splitting data

You may have noticed above that we computed the mean flipper length across all species of penguins combined. But do all species have the same mean? A common pattern in R is called "split-apply-combine". This pattern means: split the data into groups you're interested in, apply a function to each of those groups, and combine the results. One such function that performs this operation is called `tapply()`.

For example, to find the mean flipper length for *each* species, we could use:

```{r}
tapply(penguins$Flipper.Length..mm., penguins$Species, mean, na.rm = TRUE)
```

The basic format of the `tapply()` function is:

```{r}
#| eval: false
tapply(X = "data to split", INDEX = "what to split by", FUN = "what to compute")
```

Splitting can also be applied directly to `data.frame`s using the `split()` function. Let's say we wanted to split the penguins `data.frame` into separate `data.frame`s for each species.

```{r}
by_species <- split(penguins, f = penguins$Species)
```

This function returns a list of `data.frame`s, one for each species in the original data. Use `names(by_species)` to see what each list element is named. To extract the first `data.frame` from this list, which contains Adelie penguin data only, we can subset the list.

```{r}
adelie <- by_species[[1]]
```

Performing operations on lists is such a common task in R that a function exists specifically to apply functions to list elements. This function is called `lapply()`. `lapply()` takes a list and a function, applies that function to each list element, and returns the results as a new list. We'll learn more about this later in the functional programming section.

For example, to see how many rows are in each of the `data.frame`s in the `by_species` list:

```{r}
lapply(by_species, nrow)
```

### Plotting data

Data cleaning and data visualization go hand-in-hand. To effectively clean data, you should be examining the changes you're making in real-time. Base R actually has very powerful graphics capabilities for quickly visualizing data. Packages like [ggplot2](https://ggplot2.tidyverse.org/) and [lattice](https://lattice.r-forge.r-project.org/) provide powerful alternatives to base R plots, and we'll cover `ggplot2` later. For now, base R plotting can provide all we need for exploratory analyses.

One of the most useful plots for numeric data is a histogram. Histograms bin the data and plot how many occurrences of a particular bin are present, giving you an idea of the distribution.

```{r}
hist(penguins$Flipper.Length..mm.)
```

The histogram has a few arguments we can use to adjust the plot. Use `?hist()` to see the full list. Below, we can adjust the axes to be more informative and modify the number of bins we're computing.

```{r}
hist(
  penguins$Flipper.Length..mm.,
  breaks = 30,
  main = "Flipper Length of All Palmer Penguins",
  xlab = "Flipper Length (mm)",
  ylab = "Count"
)
```

The distribution appears to be bimodal. Is this because there are different species present in this plot? We can check by applying the `hist()` function to each of the groups using the list of `data.frame`s from earlier.

```{r}
# Adelie penguins
hist(
  by_species[[1]]$Flipper.Length..mm.,
  breaks = 20,
  main = "Adelie",
  xlab = "Flipper Length (mm)"
)

# Chinstrap penguins
hist(
  by_species[[2]]$Flipper.Length..mm.,
  breaks = 20,
  main = "Chinstrap",
  xlab = "Flipper Length (mm)"
)

# Gentoo penguins
hist(
  by_species[[3]]$Flipper.Length..mm.,
  breaks = 20,
  main = "Gentoo",
  xlab = "Flipper Length (mm)"
)
```

Histograms are useful for plotting the distribution of a single numeric variable. Often, we wish to see how two variables are related. The `plot()` function provides a way to create a scatter plot of two variables against each other.

```{r}
plot(
  x = penguins$Culmen.Length..mm.,
  y = penguins$Culmen.Depth..mm.,
  xlab = "Culmen Length (mm)",
  ylab = "Culmen Depth (mm)",
  main = "Relationship between culmen length and depth"
)
```

Again, these data points seem to be split by the different species of penguins present in the dataset. We can color these points using an additional variable in the call to the `plot()` function.

```{r}
plot(
  x = penguins$Culmen.Length..mm.,
  y = penguins$Culmen.Depth..mm.,
  col = factor(penguins$Species),
  xlab = "Culmen Length (mm)",
  ylab = "Culmen Depth (mm)",
  main = "Relationship between culmen length and depth"
)
```

You may have noticed that we did something special to that vector of species. We wrapped it in the `factor()` function. In R, factors are used when we have categorical values. R uses factors to represent the different levels of each category. It may not seem important now, but factors are very useful for statistical analysis and plotting. We'll build on this topic shortly.

The `plot()` function is very versatile. You can use it to create line plots as well to show trends of variables over time. Here is a contrived example of using `plot()` to create a simple line chart.

```{r}
plot(
  x = adelie$Sample.Number,
  y = adelie$Body.Mass..g.,
  type = "l",
  main = "Body mass vs sample number in Adelie penguins",
  xlab = "Sample Number",
  ylab = "Body Mass (g)"
)
```

Scatter plots display two numeric values against each other, but you'll often want to compare numeric values across categorical values. A boxplot is perfect for this. Boxplots show the median and interquartile range (IQR) for a set of data.

```{r}
boxplot(
  Flipper.Length..mm. ~ Species,
  data = penguins,
  main = "Flipper length by species"
)
```

The `boxplot()` function is a little funny because it uses formula notation. Many functions in R accept formula notation as input. In this case, `Flipper.Length..mm. ~ Species` means "plot the flipper length as a function of the species type". You may have also noticed that we gave this function the `penguins` `data.frame` via the `data =` argument. This allows the function to look up the column names without us having to explicitly type `penguins$` every time.

If there's not too much data, boxplots can actually hide a lot of information. In this case, another option is to use a stripchart, which shows every data point on the plot.

```{r}
stripchart(
  Flipper.Length..mm. ~ Species,
  data = penguins,
  method = "jitter",
  pch = 1,
  main = "Flipper length by species",
  vertical = TRUE
)
```

Bar charts are another common way to show summaries of data. To create a barplot of the mean flipper length for each species, we first need to summarize the flipper length for each group and then supply these summaries to `barplot()`.

```{r}
species_means <- tapply(
  penguins$Flipper.Length..mm.,
  penguins$Species,
  mean,
  na.rm = TRUE
)

barplot(species_means, main = "Mean flipper length by species")
```

There are many ways to modify these default plot types and make the charts look great. Take the time to look into some of base R's graphing capabilities. A really good tutorial for learning about base R plotting can be found [here](http://karolis.koncevicius.lt/posts/r_base_plotting_without_wrappers/). Also, check out `?pairs()`, `?dotchart()`, and `?coplot()` for some other useful base R graphics functions.

### Filtering data

We've already learned how to subset a `data.frame`, but let's quickly revisit it. To subset a `data.frame` means to select rows based on some condition. This is accomplished using the `[` operator and the `df[rows we want, cols we want]` syntax.

For example, in the above plots, it looked like the Gentoo penguins typically had the largest flippers. It looked like a value > 205 roughly separated the Gentoo penguins from the others. How could we select the rows where the flipper length is greater than or equal to 205 and count the species types?

```{r}
big_flippers <- penguins[penguins$Flipper.Length..mm. >= 205, ]
table(big_flippers$Species)
```

We can combine conditions to select on multiple parameters. For example, let's select the female penguins with flippers >= 205 mm.

```{r}
big_females <- penguins[
  penguins$Sex == "FEMALE" & penguins$Flipper.Length..mm. >= 205,
]

head(big_females)
```

Another important aspect of base R `data.frame`s is that they utilize `rownames()`. Using `rownames()` is uncommon in workflows from the `tidyverse` or `data.table`. However, `rownames()` can be very useful when dealing with bioinformatics data. One R package that utilizes `rownames()` extensively is [SummarizedExperiment](https://www.bioconductor.org/packages/devel/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html), which uses row and column names to ensure that data stays coordinated during an analysis.

You can view and set the rownames on a `data.frame` using the `rownames()` function.

```{r}
# View the current rownames
head(rownames(penguins))

# Set the rownames based on a new value
# paste() concatenates strings together
rownames(penguins) <- paste("row_number", 1:nrow(penguins), sep = ".")

# View the new rownames
head(rownames(penguins))
```

Above, we filtered by a column in the `data.frame`. We can also filter by selecting specific rownames.

```{r}
penguins[c("row_number.1", "row_number.5"), ]
```

### Reordering data

We can also change the order of the rows in a `data.frame`. This is similar to subsetting but instead rearranges the data based on some condition. To reorder, we use the `order()` function.

For example, we can order the data by decreasing flipper length:

```{r}
head(penguins[order(penguins$Flipper.Length..mm., decreasing = TRUE), ])
```

R also provides the `sort()` function. See if you can understand the difference between `order()` and `sort()` by looking up their help pages!

Similar to filtering above, we can also reorder by the `rownames()`.

```{r}
head(penguins[order(rownames(penguins)), ])
```

*(Can you figure out why the `data.frame` was sorted this way? Hint: Rownames are characters, so "row_number.10" comes before "row_number.2" alphabetically!)*

### Selecting columns of data

Similarly to how we select rows, we can also select columns. First, to see what columns are in our `data.frame`, we use the `colnames()` function.

```{r}
colnames(penguins)
```

To select columns, we can either specify them by name:

```{r}
head(penguins[, c("studyName", "Species")])
```

Or by their numeric index:

```{r}
head(penguins[, c(1, 3)])
```

If we want to rename a column, we can use the `colnames()` function to reassign a specific position with a new name.

```{r}
# This returns the numeric index of the colname that matches "Species"
which_col_to_change <- which(colnames(penguins) == "Species")

# Rename "Species" to "species"
colnames(penguins)[which_col_to_change] <- "species"
```

One tricky aspect of base R subsetting has to do with selecting a *single* column. If you select a single column using bracket notation, R simplifies it and returns a vector. To ensure that you return a `data.frame`, add the `drop = FALSE` argument.

```{r}
# Returns a vector
class(penguins[, "species"])

# Returns a data.frame
class(penguins[, "species", drop = FALSE])
```

### Modifying columns of data

Adding or removing columns is as simple as specifying the name of the new column using the `$` operator and assigning data to it.

For example, if I wanted to convert the flipper length from mm to m, I could multiply each value by 0.001.

```{r}
penguins$flipper_length_m <- penguins$Flipper.Length..mm. * 0.001
```

This data transformation works seamlessly because of R's *vectorization* rule described earlier. What if I wanted to define a new variable based on a logical condition? A useful function for performing this action is `ifelse()`. `ifelse()` evaluates a logical condition and returns one value if TRUE, and another if FALSE.

```{r}
penguins$is_adelie <- ifelse(
  penguins$species == "Adelie Penguin (Pygoscelis adeliae)",
  "Adelie",
  "Other"
)
```

If I want to remove a column from the `data.frame`, I can simply set its value to `NULL`.

```{r}
penguins$Flipper.Length..mm. <- NULL
```

## String operations

String operations are an important part of data cleaning. Base R supports many functions for transforming strings. Again, these functions are typically vectorized, meaning they can easily be used to modify entire columns at once.

The "species" column of the penguins `data.frame` is annoying to work with. We can simplify this column by extracting just the core species name and dropping the scientific name.

One base R function we can use to find and replace text is `gsub()`. `gsub()` takes a [regular expression](https://learn.microsoft.com/en-us/dotnet/standard/base-types/regular-expression-language-quick-reference) to find text inside of a string and replaces it with the text you supply. Regular expressions can become incredibly complex, but taking some time to understand the basics will go a long way during data cleaning.

```{r}
# Replace a space followed by the word 'penguin' and any other character with nothing ("")
penguins$species_clean <- gsub(
  pattern = " penguin.*",
  replacement = "",
  x = penguins$species,
  ignore.case = TRUE
)

# Show the counts for these new, cleaner categories
table(penguins$species_clean)
```

Another base R function that is useful for extracting text is `substr()`. `substr()` extracts sub-strings from the supplied text based on numeric start and stop positions.

```{r}
penguins$species_simple <- substr(penguins$species, start = 1, stop = 1)
table(penguins$species_simple)
```

A slightly more complicated string operation involves finding and extracting a specific pattern. This involves two functions: `regexpr()` to find the match, and `regmatches()` to extract it. Let's extract the word 'penguin' from each of the original species strings.

```{r}
penguins$only_penguin <- regmatches(
  x = penguins$species,
  m = regexpr(pattern = "penguin", text = penguins$species, ignore.case = TRUE)
)
```

R also supports basic operations for changing case. For example, to ensure `only_penguin` contains 'penguin' entirely in lowercase:

```{r}
# lower case
head(tolower(penguins$only_penguin))

# or upper case
head(toupper(penguins$only_penguin))
```

String splitting in base R can be accomplished using `strsplit()`. This function is a little tricky because it returns a *list* of split strings.

```{r}
split_species <- strsplit(
  x = penguins$species,
  split = " "
)

# strsplit returns a list of character vectors split on every space
head(split_species)

# To extract just the first word from each split list element, we can use sapply
# The `[[` tells sapply to apply the double-bracket extraction, and `1` means get the first element.
penguins$extracted_species <- sapply(split_species, `[[`, 1)
```

## Control flow

R contains all of the typical control flow mechanisms you expect from a programming language.

For-loops are somewhat under-utilized by many R users but can often be the clearest approach for new (and experienced) users. One common for-loop idiom in R is to use it in conjunction with the `seq_along()` function. `seq_along()` generates an index of the given vector that you can safely loop over.

```{r}
#| eval: false
for (i in seq_along(x)) {
  doSomething(x[i])
}
```

Other looping operations in base R include `while` and `repeat`.

Conditional statements can be tested using `if` statements. These follow a familiar syntax:

```{r}
x <- 10
if (x == 10) {
  print("the value of x is 10")
} else {
  print("the value of x is something else")
}
```

Standard `if` statements in R are *not* vectorized. Use `ifelse()` to perform a vectorized conditional statement on a column, as demonstrated earlier.

`if` statements are often combined with for-loops.

```{r}
# Make some space to save the result
result <- vector("character", length = 10)

# Determine if the number is even or odd
for (i in 1:10) {
  if (i %% 2 == 0) {
    result[i] <- "even"
  } else {
    result[i] <- "odd"
  }
}

result
```

R also contains a `switch()` statement which can be very useful when programming new functions with multiple potential inputs.

## Functions

Eventually, you'll want to create your own functions to do stuff. Functions can be defined in R using the following syntax:

```{r}
#| eval: false
f <- function(args) {
  expr
}
```

To make this a little more concrete, let's define a function that classifies a penguin based on its flipper size. Our function will take in a flipper size and a threshold value for determining whether the penguin is "large" or "small". The function will return the resulting size designation.

```{r}
determine_size <- function(flipper_size, threshold = 0.2) {
  if (flipper_size >= threshold) {
    size_category <- "large"
  } else {
    size_category <- "small"
  }

  return(size_category)
}

# Is a 0.5 meter flipper length large or small?
determine_size(flipper_size = 0.5)
```

## Functional programming

For-loops in R are simple to understand, but most R programmers heavily utilize functions in the `apply()` family. These functions supply abstractions over common for-loops that make applying functions over lists and matrices much easier in R.

The most commonly used functional programming paradigm in R is the `lapply()` function. `lapply()` takes a list as input, applies a function to each element of that list, and returns the results as a list. For example, earlier we split the penguins `data.frame` into a list of smaller `data.frame`s for each species. Instead of writing a loop to create a histogram for each one, let's define a function and `lapply()` it.

```{r}
# Define a function for plotting a histogram of flipper lengths
plot_histogram <- function(df) {
  hist(df[, "Flipper.Length..mm."], breaks = 20)
}

# Apply this function to every data.frame
# invisible() is used to suppress the text output from the hist() function
invisible(
  lapply(X = by_species, FUN = plot_histogram)
)
```

R includes several other specialized functions for functional programming. Be sure to explore `apply()` for computing over rows/columns of matrices, `sapply()` and `vapply()` for applying functions over lists and returning simplified vectors, and `mapply()` for applying a function to multiple list arguments at once.

## R packages

Even though the base language provides much of the functionality we need for simple analyses the real strength of R comes from its rich package repository called the [Comprehensive R Archive Network](https://cran.r-project.org/) or CRAN. CRAN packages can be installed using `install.packages("package_name")`. in addition to CRAN the bioinformatics community is centered around the [BioConductor](https://bioconductor.org/) project. BioConductor curates R packages focused on biological data analysis and uses standardized objects for R analysis that enable really great cross-package functionality. BioConductor packages can typically be installed using the `BiocManager` package with `BiocManager::install("bioc_package")` 

## Resources

This only scratches the surface of programming in R. Check out these resources for more information:

* [fasteR](https://github.com/matloff/fasteR)
* [R for Data Science](https://r4ds.had.co.nz/)
* [Advanced R 2nd Edition](https://adv-r.hadley.nz/)
* [Advanced R 1st Edition](http://adv-r.had.co.nz/)
* [The R Inferno](https://www.burns-stat.com/pages/Tutor/R_inferno.pdf)
